Statistical Inference of Transport Mechanisms and Long Time Scale Behavior from Time Series of Solute Trajectories in Nanostructured Membranes.

### 2020-04-08 Ben_Manuscripts/hdphmm/Outline.tex
New updates to outline

  We describe the dynamics of each state visited by solutes using a first order vector 
  autoregressive (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
        %MRS2: mean is zero in Z, is it necessarily in R (I realize we've talked about this before, but wanted to get the clarification in text)
    %BJC2: The mean I'm talking about below is different than you are describing above.
    % The multivariate Gausian noise is coming from N(0, \Sigma)
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item This means that we only parameterize $A_1$. To simplify notation, we will just
  	call it $A$.
  	\item We used a matrix-normal inverse-Wishart prior on parameters $A$ and $\Sigma$ 
  	and a Gaussian prior on $\mathbf{c}$ in order to infer their values.~\cite{fox_nonparametric_2009}
  	%BJC2: I'm not sure above is enough to make it clear how I got the parameters. Maybe I should
  	% explicitly refer to the fox paper above which outlines the inference procedure in more detail?
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %BJC3: I got the software from this page: https://homes.cs.washington.edu/~ebfox/software/
    % Maybe I should reach out to emily fox to see how she would prefer I cite the software?
    %MRS4: yes, she how she wants to cite it.
    %MRS3: be a bit more specific about what the criteria were for convergence.
    %BJC3: See if that's better
    %MRS4: But what quantitative criteria did you use to determine convergence?
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    of each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of 
    Equation~\ref{eqn:var}.
    \item Plots illustrating parameter convergence are available in the SI.
    \item We refer the interested reader to much more extensive descriptions of 
    the inference and sampling procedures used to estimate the VAR(1) parameters
    and the state sequence. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  % BJC3: I tried again with this paragraph above. I want to set up the idea that we
  % are performing this somewhat complex procedure for a good reason before jumping
  % straight into it.
  %MRS3: be more specific - i.e. describe what are the ways you are going to apply it. 
%  There are many ways one can apply the IHMM algorithm to timeseries data. 
%  \begin{itemize}
%    %MRS3: it's not as many as possible, but identify one's supported by the data, right?
%    \item We begin by trying to identify as many distinct states as possible
%    in each solute trajectory and then take advantage of the system's cylindrical
%    symmetry in order to cluster the parameter sets into an interpretable number
%    of states.
%    %MRS3: not clear why this is the case?
%    \item Having a large number of states gives more reliable clustering results.
%%    \item We believe the subsequent procedure is best suited for our data
%%    because it can clearly distinguish different dynamical behavior, it takes
%%    advantage of the system's cylindrical symmetry and represents distinct
%%    dynamics in an interpretable number of states.
%  \end{itemize}

  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry.
  \begin{itemize}
%    \item In general we are interested in the time series created by tracing
%    each solute's center of mass position. % ALready said this above
    \item There are a number of viable ways in which one could choose to analyze 
    these time series.
    \item We could use all 3 dimensions straight from the output trajectory.
    \item We could use just the $z$ dimensional coordinate since we are primarily
    interested in through-plane membrane transport.
    \item We chose to use a multi-step procedure that we believe adequately
    distinguishes every type of distinct dynamical behavior exhibited by 
    the solutes.
    \item We work in cylindrical coordinates since each pore is cylindrical
    and we expect solutes to exhibit radially symmetric dynamics.
  \end{itemize}

  We start by applying the IHMM algorithm to 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    \item By working in cartesian coordinates, we avoid mathematical complexity
    introduced by cylindrical coordinates while estimating the state sequence.
  \end{itemize}

  \begin{figure}
  \centering
  %MRS3: does this preserve the z-axis symmmetry? Is it just ensuring
  %that you can't jump that far, or is it limitinig the values that
  %the mean can take? Apologies if it's not clear to me yet.  I'm
  %pretty sure it's just limiting the jump (since y_t = c + f(y_t-1)),
  %but would be good to make that explicit - ``i.e. we need to make sure it doesn't jump too far'')
  \includegraphics[width=0.2\textwidth]{cartesian_cylinder.pdf}
  \caption{Using the nearest pore center as the origin, we represented the solute's
  location along the pore axis in terms of the $z$ coordinates and their radial distance
  from the pore centers in 2 dimensions, $x$ and $y$.}\label{fig:cartesian_cylinder}
  \end{figure}

  We applied the IHMM to each of the 24 solute trajectories independently.
  \begin{itemize}
    \item Although the IHMM is capable of identifying an infinite number of states, 
    a Dirichlet Process tends to exhibit a ``rich get richer" effect, favoring
    a fewer number of states.
    \item By applying the algorithm to each trajectory independently, we reduce
    the possibility of lumping together multiple similar states which we
    would prefer to stay separated before clustering.
%    \item From a practical implementation standpoint, the algorithm is written
%    such that we must put a cap on the maximum number of states. We chose to 
%    allow up to 100 states for each trajectory although the algorithm typically
%    finds within the range of 5-15 states. 
    % BJC2: There will probably be a figure telling the number of states for 
    % each trajectory.
  \end{itemize}

  The states identified by the IHMM are heavily influenced by the Gaussian prior
  placed on $\mathbf{c}$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of $A$ and $\Sigma$ do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the mean level of 
    each state lies within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to lie halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the
    maximum and minimum to be 2 standard deviations from the prior mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  %BJC2: More labels might be useful if not cluttered
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  %MRS4: remind people what $c$ physically represents here. 
  \caption{The parameters of the prior on the mean vector, $c$, (black line) should be chosen such
  that the mean levels of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We chose 
  $\sigma$ of the prior by defining the maximum and minimum to be 2 standard deviations
  from the mean (dashed lines).}\label{fig:prior_guesses}
  \end{figure}

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    %MRS3: can you say more which categories converged quickly?
    %BJC3: I imagine states with less sampling converge slower. But I'll need to
    % look further into this.
   \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We defined the finalized parameters of each state as the average
    of the parameters from each iteration recorded after the equilibration 
    time point.
    % MRS3: averaged and used this as estimate for running models later? 
    % BJC3: yes, well for clustering. Tried to clarified above
    % MRS3: other possbility is maximum a posteriori estimate, but for peaked 
    % symmetric distributions, it's essentially the same.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item We found the equilibration time point of each component of the 
    $\mathbf{c}$ vector as well as $A$ and $\Sigma$ matrices, and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

@@

  We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different, especially in the axial direction where solutes could
   	gßet trapped along a broad and continuous range of $z$ coordinates. 
  %MRS4: Maybe better to phrase above in terms of the symmetry in z. Why does it matter the range is continuous?
  \end{itemize}

@@

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on the two diagonal entries of $A$ and the two diagonal entries 
  of $\Sigma$.
  \begin{itemize}
   \item One can choose alternative clustering features, such as the eigenvalues of
   the $A$ and $\Sigma$ matrices, but we chose their diagonals because they can be
   easily translated to solute behavior.
   %MRS4: suggest translating them here into solute behavior for the reader!
   \item We also chose to cluster the diagonals of each matrix independently before
   combining the clusters to assign final state labels. 
   \item If we divide the total states into $m$ clusters based on the diagonal entries
   of $A$ and, separately, $n$ clusters based on the diagonal entries of $\Sigma$,
   then there are a total of $mn$ possible cluster combinations.
   \item This allows slightly more flexibility in the total number of states compared
   to clustering on all four features at once, since there can be up to, but not 
   necessarily exactly, $mn$ states. 
   %MRS4: so no clustering based on transition probablity, correct?
   %BJC4: not at the moment. That might have been more helpful if I was first-order differencing.
  \end{itemize}

  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines.
  \begin{itemize}
  	%BJC3: not sure if the commented point worth making.
%    \item We should expect a relatively high number of states. If we only consider
%    the diagonal entries of the $A$ and $\Sigma$ matrices, a total of four independent
%    variables, we can consider clustering each into at least two groups (higher versus
%    lower values), which implies $2^4$ distinct dynamical behaviors. 
%MRS5: should be lower-case 's' on silhouette
    \item We used the Silhouette test in order to score the quality of
    clusters as a function of the number of clusters chosen.
    \item For our data, the Silhouette test generally favors the lowest number
    of clusters possible (see Supporting Information).
    \item However, choosing too few clusters tends to result in 
    inadequate distinguishment between visually obvious differences in 
    dynamic behavior.
    \item This results in finalized parameter sets that are averages of
    distinct behavior which presents further problems with the predictive
    modeling that we discuss later on.
    \item We aimed to maintain the highest Silhouette score, and thus lowest
    number of clusters, possible while verifying that visually distinct 
    states stayed separated.
    \item Based on these guidelines, we decided to group the $A$ and $\Sigma$
    parameters into five clusters each. 
	\item Of the 25 distinct states allowed by this formulation, the solutes
	in this study clustered into 13--18 distinct clusters.
  \end{itemize}

  We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states compiled across all 
   trajectories.
  \end{itemize}

  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
  each value of $\mathbf{c}$ assigned to the same cluster.
  \begin{itemize}
   \item We only care about the $r$ component of $\mathbf{c}$ because solute
   trajectories are not bound in the $z$ direction.
  \end{itemize}

  We used the parameter inference component of the IHMM algorithm in order to 
  infer $A$ and $\Sigma$ of the clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segement of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    \item We drew state sequences with transition probabilities given by $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. 
  \begin{itemize}  
  	\item Our application of the IHMM does a good job of identifying and
  	distinguishing these behaviors.   	
  	\item The following approach to analysis demonstrates how one can break 
  	down the very complex behavior exhibited by solutes over long MD trajectories 
  	into discrete categories.
%  	\item Clustering the behaviors helps us to tell a coherent story which
%  	emphasizes those interactions which most heavily influence solute transport.
	\item Further, we can use the IHMM in order to generate stochastic trajectories,
	with dynamics similar to MD, so that we can estimate macroscopic transport
	properties.
  \end{itemize}

@@

  \noindent Methanol exhbits a range of dynamic behavior.
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajctory exhibits 8 distinct dynamical behaviors (in order of their first appearance,
    they are colored dark yellow, red, light blue, light green, orange, blue, teal and lime green). 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    trajectory frame.
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as to what part of
    the monomer it is hydrogen bonding, and finally, the local number density 
    surrounding the solute. 
    %and the radial location of the solute with respect to the nearest pore
    %center (this information is redundant but can be a useful visual aid).
  \end{itemize}

@@

  \noindent It is clear that there is a wealth of information in these plots. 
  \begin{itemize}
    \item One can analyze each segment sequentially and form a clear picture
    of this instance of methanol's behavior without the need to visulize
    the traejectory with software like VMD.
    \item However, a more general understanding of methanol's average behavior
    is necessary in order to formulate LLC membrane design principles.
  \end{itemize}

>>>>

  We describe the dynamics of each state visited by solutes using a first order vector 
  autoregressive (VAR(1)) model. 
  \begin{itemize}
  	\item In general, a VAR($r$) process is characterized by a vector of observations in a time series 
  	that are linearly dependent on $r$ previous values of the time series vector:
  	\begin{equation}
  	\mathbf{y}_t = \mathbf{c} + \sum_{i=1}^r A_i\mathbf{y}_{t-i} + \mathbf{e}_t~~~~\mathbf{e}_t \sim \mathcal{N}(0, \Sigma)
  	\label{eqn:var}
  	\end{equation}
  	Previous observations are weighted by coefficient matrices, $A_i$. The VAR($r$) 
  	process is further characterized by a shift in the mean of each dimension by the
  	vector $\mathbf{c}$ and a white noise term $\mathbf{e}_t$.~\cite{hamilton_time_1994}
  	\item We assumed $\mathbf{e}_t$ to be multivariate Gaussian noise, with mean zero and
  	covariance, $\Sigma$.
  	\item We limited our analysis to an autoregressive order of $r=1$.
  	\item This means that we only parameterize $A_1$. To simplify notation, we will just
  	call it $A$.
  	\item We used a matrix-normal inverse-Wishart prior on parameters $A$ and $\Sigma$ 
  	and a Gaussian prior on $\mathbf{c}$ in order to infer their values.~\cite{fox_nonparametric_2009}
  	%BJC2: I'm not sure above is enough to make it clear how I got the parameters. Maybe I should
  	% explicitly refer to the fox paper above which outlines the inference procedure in more detail?
  \end{itemize}

@@

  %Based on the algorithms developed by Fox et al.~\cite{fox_sticky_2007}, 
  Using the IHMM framework, we estimated the most likely number and sequence of hidden
  states in each solute center-of-mass trajectory while simultaneously inferring VAR(1)
  parameters for each state and the overall state transition probability matrix, $T$.
  \begin{itemize}
    \item We created a python implementation of this process which we heavily adapted from
    the MATLAB code of Fox et al.~\cite{fox_sticky_2007} 
    %BJC3: I got the software from this page: https://homes.cs.washington.edu/~ebfox/software/
    % Maybe I should reach out to emily fox to see how she would prefer I cite the software?
    %MRS4: yes, she how she wants to cite it.
    \item Parameter estimation is iterative. Therefore, we looked for convergence 
    of each entry in the $\mathbf{c}$ vector as well as the $A$ and $\Sigma$ matrices of 
    Equation~\ref{eqn:var}.
    \item We detected equilibration of the parameters using the module 
    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
    parameter estimates. 
    \item Plots illustrating parameter convergence are available in the SI.
    \item We refer the interested reader to much more extensive descriptions of 
    the inference and sampling procedures used to estimate the VAR(1) parameters
    and the state sequence. 
    ~\cite{beal_infinite_2002,teh_hierarchical_2006,van_gael_beam_2008,fox_nonparametric_2009,fox_bayesian_2010}
  \end{itemize}

  We carefully applied the IHMM algorithm in a way which takes advantage of the
  system's cylindrical symmetry.
  \begin{itemize}
%    \item In general we are interested in the time series created by tracing
%    each solute's center of mass position. % ALready said this above
    \item There are a number of viable ways in which one could choose to analyze 
    these time series.
    \item We could use all 3 dimensions straight from the output trajectory.
    \item We could use just the $z$ dimensional coordinate since we are primarily
    interested in through-plane membrane transport.
    \item We chose to use a multi-step procedure that we believe adequately
    distinguishes every type of distinct dynamical behavior exhibited by 
    the solutes.
    \item Ultimately, we work in cylindrical coordinates since each pore is a 
    cylinder where we expect solutes to exhibit radially symmetric dynamics.
  \end{itemize}

  We start by applying the IHMM algorithm to the 3D solute center-of-mass coordinate 
  trajectories transformed relative to the closest pore center.
  \begin{itemize}
    \item We tracked the solute's motion along the pore axis with the center-of-mass
    $z$ coordinate.
    \item Using the nearest pore center as the origin, we represented the radial 
    distance of each solute's center-of-mass from the pore center in 2 dimensions,
    $x$ and $y$ (see Figure~\ref{fig:cartesian_cylinder}).
    \item By working in cartesian coordinates, we avoid mathematical complexity
    introduced by cylindrical coordinates while estimating the state sequence.
  \end{itemize}

  \begin{figure}
  \centering
  %MRS3: does this preserve the z-axis symmmetry? Is it just ensuring
  %that you can't jump that far, or is it limitinig the values that
  %the mean can take? Apologies if it's not clear to me yet.  I'm
  %pretty sure it's just limiting the jump (since y_t = c + f(y_t-1)),
  %but would be good to make that explicit - ``i.e. we need to make sure it doesn't jump too far'')
  %BJC5: preserving z-axis symmetry isn't needed yet at this stage. I use the
  % means in z in order to distinguish states (specifically estimating the c vector).
  % I get the radial means this way. Then just ignore z later.
  \includegraphics[width=0.2\textwidth]{cartesian_cylinder.pdf}
  \caption{Using the nearest pore center as the origin, we represented the solute's
  location along the pore axis in terms of the $z$ coordinates and their radial distance
  from the pore centers in 2 dimensions, $x$ and $y$.}\label{fig:cartesian_cylinder}
  \end{figure}

  \noindent We applied the IHMM to each of the 24 solute trajectories independently.
  \begin{itemize}
    \item Although the IHMM is capable of identifying an infinite number of states, 
    a Dirichlet Process tends to exhibit a ``rich get richer" effect, favoring
    a fewer number of states.
    \item By applying the algorithm to each trajectory independently, we reduce
    the possibility of lumping together multiple similar states which we
    would prefer to stay separated before clustering.
%    \item From a practical implementation standpoint, the algorithm is written
%    such that we must put a cap on the maximum number of states. We chose to 
%    allow up to 100 states for each trajectory although the algorithm typically
%    finds within the range of 5-15 states. 
    % BJC2: There will probably be a figure telling the number of states for 
    % each trajectory.
  \end{itemize}

  % BJC5: For the paper, this paragraph and figure might be better suited for the supporting information
  % BJC5: maybe even for the thesis
  \noindent The states identified by the IHMM are heavily influenced by the Gaussian prior
  placed on $\mathbf{c}$ in Equation~\ref{eqn:var}.
  \begin{itemize}
    \item The entries of $A$ and $\Sigma$ do not vary over a wide range, so
    the final parameters were relatively insensitive to the priors.
    \item In order to maximally automate the IHMM procedure, we attempted to
    parameterize the prior on $\mathbf{c}$ in an intelligent way.
    \item The prior parameters should be chosen such that the mean level of 
    each state lies within a region of reasonable probability of the prior (see
    Figure~\ref{fig:prior_guesses}).
    \item In each dimension, we defined the prior mean to lie halfway between the 
    maximum and minimum of each trajectory dimension. 
    \item To parameterize the prior's variance in each dimension, we defined the
    maximum and minimum to be 2 standard deviations from the prior mean.
    \item Although this approach has worked quite well for the data in this work,
    it is important to check the results to determine whether further adjustments
    to the prior might be needed.
    \item In the Supporting Information, we show the result of a parameterization
    where the prior parameters of $c$ were poorly chosen. %TODO
  \end{itemize}

  %BJC2: I think it would be useful to continue using this trajectory to illustrate 
  % the parameterization process. It might be bunch of plots stacked on each other.
  %BJC2: More labels would be useful if not cluttered
  \begin{figure}
  \centering
  \includegraphics[width=\textwidth]{prior_guesses.pdf}
  \caption{The parameters of the prior on the mean vector, $c$, (black line), which
  represents the coordinates at which solute are trapped, should be chosen such
  that the mean levels of each state identified in the trajectory (blue line) lie within
  regions of the prior with reasonable probability. We chose the mean of the prior 
  as halfway between the maximum and minimum of each trajectory dimension. We chose 
  $\sigma$ of the prior by defining the maximum and minimum to be 2 standard deviations
  from the mean (dashed lines).}\label{fig:prior_guesses}
  \end{figure}

  We ran 2000 iterations of the IHMM procedure in order to arrive at converged 
  state sequences and parameters for each state.
  \begin{itemize}  
    %MRS3: can you say more which categories converged quickly?
    %BJC3: I imagine states with less sampling converge slower. But I'll need to
    % look further into this.
   \item Most parameters converged quickly (within 50-100 iterations) while
    others took up to 1000 iterations.
    \item We defined the finalized parameters of each state as the average
    of the parameters from each iteration recorded after the equilibration 
    time point.
    % MRS3: averaged and used this as estimate for running models later? 
    % BJC3: yes, well for clustering. Tried to clarified above
    % MRS3: other possbility is maximum a posteriori estimate, but for peaked 
    % symmetric distributions, it's essentially the same.
    %BJC5: moved following point to earlier in methods.
%    \item We detected equilibration of the parameters using the module 
%    \texttt{pymbar.timeseries.detect\_equilibration} on the time series of 
%    parameter estimates. 
    \item We found the equilibration time point of each component of the 
    $\mathbf{c}$ vector as well as $A$ and $\Sigma$ matrices, and then used the longest
    equilibration time of all dimensions as the equilibration time point.
  \end{itemize}

@@

  \noindent We clustered like parameter sets in order to reduce the state space to
  a more easily interpretable size.
  \begin{itemize}
  	\item For each solute studied, we identified 200-325 independent states, each
   	with separate VAR(1) parameters.
   	\item Many of these states exhibit very similar dynamical behavior except their
   	mean levels are different.
  \end{itemize}

@@

  % BJC2: this part may expand a bit if I try different clustering approaches. But
  % this is the one I'm going with for now.
  We clustered based on the two diagonal entries of $A$ and $\Sigma$ as well as the
  radial means, $\mu_r$, for a total of 5 variables.
  \begin{itemize}
   \item One can choose alternative clustering features, such as the eigenvalues of
   the $A$ and $\Sigma$ matrices, but we chose their diagonals because they can be
   easily translated to solute behavior.
   %MRS4: suggest translating them here into solute behavior for the reader!
   \item We also chose to cluster the diagonals of each matrix and the radial means
   independently before combining the clusters to assign final state labels. 
   \item If we divide the total states into $m$ clusters based on the diagonal entries
   of $A$, $n$ more clusters based on the diagonal entries of $\Sigma$, and $k$ clusters
   based on $\mu_r$, then there are a total of $mnk$ possible cluster combinations.
   \item This allows slightly more flexibility in the total number of states compared
   to clustering on all five features at once, since there can be up to, but not 
   necessarily exactly, $mnk$ states. 
   %MRS4: so no clustering based on transition probablity, correct?
   %BJC4: not at the moment. That might have been more helpful if I was first-order differencing.
  \end{itemize}

  Choosing the number of clusters can be somewhat subjective so
  we attempted to add some structure to the selection process by following
  a set of qualitative and quantitative guidelines.
  \begin{itemize}
  	%BJC3: not sure if the commented point worth making.
%    \item We should expect a relatively high number of states. If we only consider
%    the diagonal entries of the $A$ and $\Sigma$ matrices, a total of four independent
%    variables, we can consider clustering each into at least two groups (higher versus
%    lower values), which implies $2^4$ distinct dynamical behaviors. 
    \item We used the silhouette test in order to score the quality of
    clustering as a function of the number of clusters chosen.
    \item For our data, the silhouette test generally favors the lowest number
    of clusters possible (see Supporting Information).
    \item However, choosing too few clusters tends to result in 
    inadequate distinguishment between visually obvious differences in 
    dynamic behavior.
    \item This results in finalized parameter sets that are averages of
    distinct behavior which presents further problems with the predictive
    modeling that we discuss later on.
    \item We aimed to maintain the highest silhouette score, and thus lowest
    number of clusters, possible while verifying that visually distinct 
    states stayed separated.
    \item Based on these guidelines, we decided to group the $A$ and $\Sigma$
    parameters into five clusters each and the radial means into three clusters. 
	\item Of the 75 distinct states allowed by this formulation, the solutes
	in this study clustered into 30--33 distinct clusters.  %BJC5: need to check these numbers
	\item While this may seem like a high number of states, many are sampled
	infrequently.
  \end{itemize}

  \noindent We remapped the state sequence based on the cluster assignments and
  generated a state transition probability matrix, $T$.
  \begin{itemize}
   \item The IHMM algorithm also produces an estimate of $T$, but since we 
   fixed the state sequence, we decided to explicitly calculate $T$ by 
   counting the number of transitions between states compiled across all 
   trajectories.
  \end{itemize}

%  We obtained $\mathbf{c}$ vectors of the clustered states by averaging 
%  each value of $\mathbf{c}$ assigned to the same cluster.
%  \begin{itemize}
%   \item We only care about the $r$ component of $\mathbf{c}$ because solute
%   trajectories are not bound in the $z$ direction.
%  \end{itemize}

  We used the parameter inference component of the IHMM algorithm in order to 
  infer $A$, $\Sigma$ and $\mu_r$ of the clustered states.
  \begin{itemize}
   %BJC2: I'm unsure of the precise reasoning here.
   \item We could not simply take the mean of the clustered 
   $A$ and $\mathbf{e}_t$ parameters because it is not clear that this is a linear 
   operation for this problem. 
   \item To circumvent this problem, we modified the ($r$, $z$) solute
   trajectories so that they had a mean of zero, leaving only the 
   fluctations. 
   \item We did this by subtracting the maximum likelihood estimate
   of the $\mathbf{c}$ vector from each same-state segment of the unclustered
   trajectory.   
   \item We used the IHMM algorithm on this modified trajectory to infer
   the clustered state parameters by fixing the clustered state sequence. 
  \end{itemize}

  \noindent Finally, we generated stochastic trajectory realizations using the clustered
  parameter sets.
  \begin{itemize}
    \item We drew state sequences with transition probabilities given by $T$.
    \item While in a given state, we simulated motion according to the VAR(1)
    parameterization of that state.
    \item After each state transition, we reset the unconditional mean of each 
    state based on the particle's position immediately before the state transition
    occurred.
  \end{itemize}

  % BJC5: not sure what to title this. Or break it into more sections.
  \subsection{Calculation of Physical Interactions}\label{method:interactions}

  We determined the number of hydrogen bonds between each solute and the membrane
  as a function of time.
  \begin{itemize}
    \item Based on the geometric criteria of Luzar and Chandler, we define a hydrogen
    bond to exist if the distance between donor, D, and acceptor, A, atoms is less 
    than 3.5\AA~and the angle formed by $D-H \cdots A$ is less than 30
    $\degree$.~\cite{luzar_effect_1996}
  \end{itemize}

  We estimated the lifetime of hydrogen bonds by recording the length of 
  sequential frames where solute remained hydrogen bonded.
  \begin{itemize}
    \item We still counted sequences where hydrogen bonds were broken for
    a single frame before reforming.
    \item We reported the 95th percentile of hydrogen bond lifetimes.
  \end{itemize}

  We also measured the degree of association between solutes and sodium ions on a
  frame-by-frame basis.
  \begin{itemize}
    \item We define a sodium ion to be associated with an atom if they are within 
    2.5 \AA~of each other, as determined in our previous work.~\cite{coscia_chemically_2019}
  \end{itemize}

  We developed a way to measure the local density of the membrane at arbitrary
  points in the unit cell.
  \begin{itemize}
    \item We histogrammed the positions in three dimensions
    \item Since our system is in a monoclinic unit cell, we periodically replicated
    the system in +/- x, y and z direction and then set the bounds on the
    histogram to create a rectanglular box encompassing the unitcell with a 1 nm
    buffer between the histogram unitcell boundaries.
    \item We then used a regular grid interpolator in order to allow interpolation
    at arbitray points within the grid.
  \end{itemize}

  We quantified average solute movement using the time averaged mean squared 
  displacment (MSD) of the solute center of mass trajectories.
  \begin{itemize}

 	\item The time-averaged MSD measures all observed displacements over time lag $\tau$:
  	\begin{equation}
  	\overline{z^2(\tau)} = \dfrac{1}{T - \tau}\int_{0}^{T - \tau} (z(t + \tau) - z(t))^2 dt
  	\label{eqn:tamsd}
  	\end{equation}
  	where T is the length of the trajectory.
  	\item We reported 1$\sigma$ confidence intervals based on the results of 200 
  	bootstrap trials where calculated the mean of $n$ randomly chosen trajectories 
  	with replacement from the pool of trajectories, where $n$ is the number of trajectories.
  \end{itemize}

@@

  Solutes in this system show a wide range of behavior influenced by the 
  heterogeneity of the membrane's nanostructure as well as the interactions 
  between monomer and solute chemical functionality. 
  \begin{itemize}  
  	\item Our application of the IHMM does a good job of identifying and
  	distinguishing these behaviors.   	
  	\item The following approach to analysis demonstrates how one can discover and explain the very complex behavior exhibited by solutes over long MD trajectories 
   in terms of solute-membrane interactions.
%  	\item Clustering the behaviors helps us to tell a coherent story which
%  	emphasizes those interactions which most heavily influence solute transport.
	\item Further, we can use the IHMM in order to generate stochastic trajectories,
	with dynamics similar to MD, so that we can estimate macroscopic transport
	properties.
  \end{itemize}

@@

  \noindent Methanol exhbits a range of dynamic behavior.
  \begin{itemize}
    \item In Figure~\ref{fig:mechanism_map}, we plot a representative 2D methanol 
    trajectory in cylindrical coordinates.
    \item The coordinates are colored according to the cluster with which methanol's
    dynamics are most consistent, which implies that this particular methanol 
    trajectory exhibits 8 distinct dynamical behaviors (in order of their first appearance,
    they are colored dark yellow, red, light blue, light green, orange, blue, teal and lime green). 
    \item The solute timeseries are coupled with color coded bars which describe the
    solute's local environment and physical interactions that it undergoes at that
    trajectory frame.
    \item From the top, they describe whether the solute is associated with sodium,
    whether the solute is hydrogen bonding with the monomer as well as to what part of
    the monomer it is hydrogen bonding, and finally, the local number density 
    surrounding the solute. 
    %and the radial location of the solute with respect to the nearest pore
    %center (this information is redundant but can be a useful visual aid).
  \end{itemize}

@@

  \noindent It is clear that there is a wealth of information in these plots, but they don't give the complete picture needed in order to formulate LLC membrane design principles.
  \begin{itemize}
    \item One can analyze each segment sequentially and form a clear picture
    of this instance of methanol's behavior without the need to visulize
    the traejectory with software like VMD.
    \item However, it is still possible to gain a more general understanding of methanol's average behavior.
  \end{itemize}
